---
title: "Dietary Choline Intake Adequacy and Liver Enzymes by BMI in U.S. Adults: A Cross-Sectional Analysis of NHANES 2021-2023"
author:
  - name: Alfonso L Neri
    orcid: 0000-0001-7004-5464
    corresponding: true
    email: alneri@uab.edu
    roles:
      - Conceptualization
      - Methodology
      - Formal analysis
      - Software
      - Visualization
      - Writing - original draft
    affiliations:
      - ref: uab-hpo
  - name: Theodoros Giannouchos
    orcid: 0000-0002-1574-6767
    corresponding: false
    roles:
      - Supervision
      - Methodology
      - Writing - review & editing
    affiliations:
      - ref: uab-hpo
affiliations:
  - id: uab-hpo
    name: University of Alabama at Birmingham
    department: Department of Health Policy and Organization, School of Public Health
    city: Birmingham
    region: AL
    country: USA
keywords:
  - choline
  - Adequate Intake
  - obesity
  - body mass index
  - liver enzymes
  - dietary reference intakes
  - NHANES
abstract: |
  **Background:**

  **Aim:**

  **Methods:**

  **Results:** 
  
  **Conclusion:** 
date: last-modified
format:
  html: default
  pdf: default
  docx: default
bibliography: references.bib
csl: sage-harvard.csl
number-sections: false
---

# Introduction

```{r}
#| label: setup
#| include: false
#| message: false

library(gt)
library(gtsummary)
```

The Adequate Intake (AI) for choline was established based on the results of controlled feeding studies in which choline was removed and then re-introduced to the diet of healthy U.S. adults [@zeisel1991; @fischer2007]. In both studies, choline deprivation resulted in sub-clinical hepatic steatosis and elevated liver enzymes that were reversed upon re-introduction of choline [@zeisel1991; @fischer2007]. Because these data were limited in number, the Institute of Medicine could not establish an Estimated Average Requirement (EAR), and thus the AI was established in 1998 to be 550 mg/day for men and 425 mg/day for women, which has not been updated [@instituteofmedicine1998].

It is estimated that a substantial proportion of U.S. adults do not meet this choline AI [@freedman2024; @wallace2018]. Since 1998, when the AI was established, obesity prevalence has more than doubled in the U.S. [@decleene2026]. This rise in obesity raises the possibility that the current choline AI may be inadequate in adults with higher BMIs.

The liver can synthesize choline via the PEMT pathway, providing approximately 1/3 of phosphatidylcholine and serving as a buffer in times of low choline intake [@vance2013]. However, this endogenous buffer mechanism may be compromised in obesity. Individuals with severe obesity have diminished liver PEMT activity [@kartsoli2020], and adults with obesity have lower serum choline than their healthy-weight counterparts [@winslow2025]. It is possible that obese individuals depend more on dietary intake to meet their choline needs, and hence impaired liver function from inadequate intake may be more severe in this population. Obesity and choline needs in obesity as well as choline intake in free-living populations have been identified as research priorities at the 2025 Choline Research Symposium [@paules2025].

Adults with obesity may be at increased risk for inadequate choline intake possibly due to dietary patterns. If inadequate choline intake is related to liver injury biomarkers in adults with obesity, the current AI may not be adequate to protect against liver injury. Recent NHANES data on choline intake and liver enzymes levels, with body mass index (BMI) adjusted as a covariate, were reported with BMI not as a focus for the examination. No study has investigated either of these possibilities in a nationally representative sample.

Thus, there were two aims of this study: 1) estimate the prevalence of inadequate choline intake among adults, stratified by BMI, sex and, among women, menopausal status; and 2) determine the relationships of inadequate choline intake and obesity with serum liver enzymes: alanine amino transferase (ALT), aspartate amino transferase (AST), and gamma-glutamyl transferase (GGT).

# Methods

We conducted a cross-sectional analysis of data from National Health and Nutrition Examination Survey (NHANES) 2021 to 2023 cycle. NHANES is conducted by the National Center for Health Statistics (NCHS) and is a complex, stratified, multistage probability sample of the civilian, non-institutionalized, US population [@nhanes_data]. NHANES participants complete a household survey as well as a physical exam, which includes body measurements and a blood draw.

The survey was approved by the NCHS Ethics Review Board, and all participants provided written informed consent. This secondary analysis of publicly available, deidentified data was not subject to ethics review. Reporting was conducted in accordance with the Strengthening the Reporting of Observational Studies in Epidemiology (STROBE) guidelines for cross-sectional studies [@von2007].

Dietary intake was assessed by two 24-hour dietary recalls using the USDA Automated Multiple-Pass Method [@moshfegh2008]. Choline and total energy (kcal/d) values were derived using the USDA Food and Nutrient Database for Dietary Studies [@usda_fndds]. Only reliable recalls from NCHS, meeting the minimum data quality standards on both days, were included in the analyses and intake was calculated as the average of the two recall days.

For aim 1, mean intake was compared to the sex-specific AI of 550 mg/d for men and 425 mg/d for women [@instituteofmedicine1998]. We classified participants as either below the AI or above the AI.

For aim 2, choline was expressed as density (mg per 1,000 kcal) and divided into quartiles within the Aim 2 sample. Density was used because absolute choline intake largely tracks total food intake. In this sample, absolute intake was moderately correlated with energy intake (r = 0.66), while density was not (r = -0.13). Expressing intake as density separates the choline content of the diet from the amount of food eaten.

In aim 2, choline intake was modeled as the density of choline intake (mg/1,000 kcal) and also divided into four quartiles in the Aim 2 sample. Choline density was used because choline intake is strongly correlated with other measures of food intake, including total energy intake (r = 0.66), in this population. In contrast, choline density was not correlated with energy intake (r = -0.13).Choline density accounts for the dietary concentration of choline and removes the influence of total food intake. In the Aim 2 sample, quartile cutoffs were 121.52, 156.53, 202.23.

BMI was calculated from measured height and weight and grouped as normal (18.5 to 24.9 kg/m²), overweight (25.0 to 29.9 kg/m²), or obese (30.0 kg/m² or greater). Underweight adults (BMI below 18.5 kg/m²) were excluded from BMI comparisons as the group was too small for stable weighted estimates.

BMI was calculated with measured height and weight and categorized as normal weight (18.5 to 24.9 kg/m²), overweight (25.0 to 29.9 kg/m²), and obese (≥(30.0 kg/m²) in both Aims. Underweight adults (BMI \<18.5 kg/m²) were not analyzed analyses because the number of cases was too low for a stable weighted estimate.

## Outcomes

The outcome in Aim 1 was inadequate choline intake (dietary intake \<AI). The outcomes for Aim 2 were serum concentrations of the liver enzymes alanine aminotransferase (ALT), aspartate aminotransferase (AST), and gamma glutamyltransferase (GGT) in the standard biochemistry panel. The values were log-transformed to reduce the skewness of the distribution to the right.

## Analyzed Sample

In adult women, menopausal status was derived from questions in the reproductive health questionnaire [@nchs_rhq_l]. Women who reported not having menstrual periods for 1 year or longer due to menopause were coded as postmenopausal, and those who still menstruated regularly were coded as premenopausal. Those who did not have periods because of hysterectomy were assigned no status (that is, "not applicable") because the public use file for this cycle does not include information to determine whether ovaries were present.

The Aim 1 sample of adults aged ≥20 years included those who provided two good-quality 24-hour dietary recalls, had a valid 2-day food weight record, and had a measured BMI. Adults with implausible estimated energy intake were excluded from analyses: those with mean intake \<500 kcal/day or \>8,000 kcal/day for men, and \<500 kcal/day or \>5,500 kcal/day for women.

The Aim 2 sample included those with a measured ALT value and those without self-reported liver disease, hepatitis B infection, and heavy alcohol use, defined as \>2 drinks/day for men and \>1 drink/day for women. Heavy alcohol use and liver disease are known to elevate liver enzymes independent of choline intake. Reported hepatitis C infection was not included in the public use file for this cycle.

All exclusions were maintained as flags rather than creating a subset of the data so that the full sample could be used to specify the survey design.

## Statistical analysis

In both aims, the two-day dietary sample weights were used together with survey stratum and primary unit identifiers, following NCHS analytic guidelines.[@nchs_analytic_guidelines]. The survey design was defined using the full dataset, and subgroups were specified by subsetting the survey design, which maintained the full variance estimation. Standard errors were computed using Taylor series linearization [@survey2010].

For Aim 1, the proportion of adults with inadequate choline intake (\<AI) was estimated using the two-day dietary sample weights for the full sample and stratified by BMI category, sex and BMI category, and for women by menopausal status and BMI category. Rao-Scott chi-square tests were used to determine if proportions varied across BMI categories.

For Aim 2, there were only 15 design degrees of freedom in this cycle, which limits the ability to estimate complex models. For example, testing for the main effects of choline quartile and BMI and their interaction would overfit the data and is therefore not pursued. Instead, a single a priori contrast was defined between adults in the lowest quartile of choline density and obese adults with all other subjects to evaluate the hypothesis that dietary inadequacy is most important when endogenous production is also low. Models were fit using the two-day dietary sample weights for each of the log-transformed liver enzymes using linear regression, adjusting for age, sex, race/ethnicity, and family income-to-poverty ratio. Regression coefficients were transformed to ratios of geometric means and 95% confidence intervals. Because there were three outcomes, Bonferroni adjustment was used to evaluate statistical significance, resulting in a required level of significance of P \< 0.017.

A secondary analysis was performed with BMI categories (normal, overweight, and obese) and a dichotomous variable for low choline intake, where the reference group was normal BMI adults with high choline intake. Six groups were formed, where each was a combination of BMI category and low-choline intake status. Joint Wald tests were performed to evaluate differences in enzymes across the six groups. A second model was fit with BMI category, low choline intake status, and their interaction to evaluate differences between low and high choline intake for each BMI category.

The code and data for the analyses is available at <https://github.com/LouieNeri/choline-bmi-nhanes.> Analyses were performed using R version 4.3.3 \[29\] using the following packages: nhanesA [@nhanesA], survey [@survey2004; @survey2024], tidyverse [@tidyverse], and gtsummary [@gtsummary].

# Results

The characteristics of the Aim 1 sample included 2,324 women and 1,836 men with a mean age of 48.9 years (SD = 17.3). The proportions with normal-weight, overweight, and obese BMI were 24.6, 33.5, and 41.9%, respectively. The mean overall choline intake was 335.0 mg/day, and means were higher for men, 384.5 mg/day, than women, 287.4 mg/day.

```{r}
#| label: tbl-baseline
#| tbl-cap: "Characteristics of the Aim 1 sample by sex, NHANES August 2021 to August 2023"
#| echo: false

readRDS("../../output/aim1_table_baseline.rds") 
```

The Aim 2 sample included 2,324 adults. Those with liver disease, hepatitis B, and high alcohol intake, and those without ALT results were excluded.

The proportion below the AI among all adults was 84.1% (95% CI:82.2-85.9 ), and the proportions below the AI did not vary across BMI categories: 82.7% (95% CI: 79.8-85.7) for normal weight, 84.9% (95% CI: 82.4-87.3) for overweight, and 84.2% (95% CI: 80.6-87.8) for obesity (p = 0.63).

Patterns varied somewhat by sex, but these differences were not significant (@tbl-aim1-sex). Among women, the proportion below the AI was 80.1% (95% CI: ) for normal weight, 85.9% (95% CI: ) for overweight, and 87.0% (95% CI: ) for obesity (p = ). In contrast, the proportions among men were 85.8% (95% CI: ) for normal weight, 84.0% (95% CI: ) for overweight, and 81.1% (95% CI: ) for obesity (p = ). The CIs overlapped for both sexes for all BMI categories.

```{r}
#| label: tbl-aim1-sex
#| tbl-cap: "Prevalence of choline intake below the AI by sex and BMI category"
#| echo: false

readRDS("../../output/aim1_table_sex_bmi.rds")
```

Among women, 680 were premenopausal, and 930 were postmenopausal, based on self-report. Among premenopausal women, the proportion of those below the AI increased with BMI, from 76.6% (95% CI: 71.8-81.4) for normal weight, to 83.1% (95% CI: 76.2-90.0) for overweight, and 88.8% (95% CI: 82.2-95.5) for obesity (@tbl-aim1-meno). In contrast, the proportions for postmenopausal women were 83.2% (95% CI: ) for normal weight, 89.0% (95% CI: ) for overweight, and 82.2% (95% CI: ) for obesity, and no clear trend was apparent. (test for BMI × menopause interaction?) Of the remaining 714 women, 393 had had a hysterectomy and 321 had missing data on their menstrual history.

```{r}
#| label: tbl-aim1-meno
#| tbl-cap: "Prevalence of choline intake below the AI among women by menopausal status and BMI category"
#| echo: false

readRDS("../../output/aim1_table_menopause.rds")
```

The geometric means of ALT and GGT for obese adults were greater than those with normal weight across all quartiles of choline intake (@tbl-enzymes). Among obese adults, the estimates were lowest for the lowest quartile of choline intake for ALT, 18.7 U/L (95% CI: 17.5-19.9), and for AST, 19.5 U/L (95% CI: 18.5-20.6), which are in the opposite direction than expected. No estimates met the suppression rule.

```{r}
#| label: tbl-enzymes
#| tbl-cap: "Survey-weighted geometric mean liver enzymes by BMI category and choline density quartile"
#| echo: false

readRDS("../../output/aim2_table_enzymes.rds")
```

Among adults, the ratio of the low choline intake and obese group to the other three groups was 1.04 (95% CI: 0.97-1.12; p = 0.26) for ALT, 0.97 (95% CI: 0.90-1.03; p = 0.27) for AST, and 1.17 (95% CI: 1.03-1.33; p = 0.019) for GGT (@tbl-contrast). These results do not reach the Bonferroni-corrected level of p \< 0.017.

```{r}
#| label: tbl-contrast
#| tbl-cap: "Adjusted ratios of geometric mean liver enzymes for adults with both low choline intake and obesity compared with all other adults"
#| echo: false

readRDS("../../output/aim2_table_contrast.rds")
```

For the secondary analysis, the sample sizes for all 6 groups were adequate (137 to 681 adults) and there was a significant difference in ALT and GGT across the 6 groups (P=0.001 for each). AST did not differ (P=0.13). The estimates for the enzymes followed the pattern for BMI category, not choline intake.

```{r}
#| label: tbl-groups
#| tbl-cap: "Adjusted ratios of geometric mean liver enzymes across six BMI and choline intake groups"
#| echo: false

readRDS("../../output/aim2_table_groups.rds") |> as_flex_table()
```

# Discussion

More than 80% of U.S. adults in this sample consumed less than the AI for choline, and the percent was not altered by BMI category. Contrary to the hypothesis, obese adults with low choline intake did not have higher ALT, AST or GGT than other adults. The difference in GGT was in the opposite direction and non-significant after correction for multiple testing.

The 84.1% of adults consuming less than the AI for choline was slightly lower than the 90% generally cited, and similar to that reported in other analyses of NHANES data [@wallace2016; @wallace2018; @freedman2024]. The lower prevalence is likely due to the method of dietary assessment. The 2-day recall used in our analysis retains more of the day-to-day variability in food intake than other methods, expanding the distribution and including more extreme values than usual intake methods [@dodd2006]. This would affect all BMI categories in the same way, so it does not explain the lack of difference between the BMI categories.

The main finding for Aim 1 was that the gradient in choline intake across BMI categories was near zero. If obese adults had consumed less choline, then the lower circulating choline in these individuals [@winslow2025] would be due to dietary intake. However, they did not consume less choline than other adults, and this was true for both absolute and density measures of choline intake. The lower circulating choline levels of obese adults [@winslow2025] is, therefore, unlikely to be due to differences in choline intake. Other mechanisms, such as differences in absorption, liver metabolism or internal production of choline, might be responsible. For example, lower expression of PEMT in the liver of adults with severe obesity [@kartsoli2020], which could not be assessed in this study, might be one explanation.

We noted an increase in the percent of premenopausal women with choline intakes below the AI with increasing BMI category. This observation might be related to the well-known role of estrogen in regulating PEMT activity [@fischer2007]. However, we emphasize that the finding is hypothesis-generating only, because of the overlapping confidence intervals, smaller cell sizes, and lack of correction for multiple testing. Notably, the trend was opposite (although small) for men, suggesting that we are unlikely to identify a single mechanism explaining the findings for the different groups.

Our negative finding for Aim 2 is consistent with a recent analysis of NHANES data that found low choline intake was not associated with liver enzymes in the U.S. population in general [@wallace2025a]. We extended that finding by examining whether obesity, which is known to lower endogenous choline synthesis, would change the nature of the association. It did not.

There are 2 possible explanations for our Aim 2 finding. First, liver enzymes, the serum aminotransferases, may be too insensitive to detect differences in chronic choline intake. For example, in a controlled feeding trial, plasma choline and betaine levels distinguished between low intake and intakes at the AI, but serum enzymes did not [@trujillo-gonzalez2026]. Second, studies that have used a direct measure of liver fat also found no association between choline intake and liver fat. For example, a cross-sectional analysis of NHANES data from 2017-2020, using FibroScan to measure liver fat, found no association between total choline intake and presence of metabolic dysfunction-associated steatoic liver disease [@taesuwan2025]. Therefore, the lack of association is not likely due to our choice of enzyme data. The lack of association is present with a more direct assessment of liver health as well. Our negative finding is more likely due to the limitations of dietary recall as a method to assess food intake in free-living adults.

Our secondary analysis also showed a significant difference in the liver enzymes for ALT and GGT. The estimates followed the pattern for BMI category, not choline intake, which is what expected based on the known associations between fat and liver enzymes.

This study had several strengths. We used the latest NHANES data and included both measured BMI and liver enzymes. We used sample weights to provide a nationally representative sample of U.S. adults. The study design was in accord with the methodology guidelines provided by the NCHS for the full sample [@nchs_analytic_guidelines]. We chose a single contrast that was pre-specified to ensure we stayed within the available degrees of freedom. All of our code was made available to the public so the results can be reproduced exactly.

A cross-sectional study design does not allow for inferences of causality. The dietary intake values reported were averaged between the 2-day recall rather than being modeled as a distribution of usual intake, which limits the ability to estimate true prevalence rates for individuals across the entire range of choline intake; however, the comparison of BMI groups is unlikely to be affected by this limitation. Individuals with higher BMI are known to underreport their intakes more frequently, which could attenuate the difference in reported intakes between the groups [@moshfegh2008]. In this study, there was a lack of degrees of freedom to allow for testing an interaction model of choline intake and alcohol intake, whereas the degrees of freedom for all 6 groups was very small with df =5 for the denominator. Heavy alcohol intake was classified based on the average number of drinks per drinking day, which may have excluded some individuals who drank moderate volumes of alcohol per day but drank less frequently, which decreased the sample size tested in Aim 2.

Additional study limitations include the inability to rule out hepatitis C without the use of clinical data, which were not available in this data cycle, and the use of BMI as a proxy for adiposity and the inability to assess liver fat content. Participants reporting a hysterectomy were not able to be placed in a menopausal status and were, therefore, excluded.

These findings do not provide data on whether or not the AI can be met with adults with obesity, despite the link between dietary recall and liver enzymes. It appears that choline intake levels are not different across BMI categories, and liver enzymes available in national population surveys do not appear to reflect habitual intake. A controlled feeding study across a broad range of BMIs would be more ideal to determine whether there are differences in choline requirements in obesity, with circulating choline and choline metabolites or direct measures of liver fat content as outcomes.
